Sampling Strategies in Decisions from Experience

### 2022-05-27 manuscript.Rmd
Submitted version (master's thesis)

authornote: |
 This manuscript is a dynamic document created from the materials provided on the `GitHub` repository: https://github.com/linushof/sampling-in-dfe. The current version of the manuscript, created from the commit with the hash ``r repro::current_hash()``, is submitted as a master's thesis in fulfillment of the requirements for the degree Master of Science (M.Sc.) at Heidelberg University, Faculty of Behavioural and Cultural Studies, Institute of Psychology. Matr.-Nr.: Removed from online version. Supervisors:

@@

abstract: |
  In the sampling paradigm, people can---as much as they want, in the way they want, and without costs---sample the possible outcomes of prospects before making a final choice. Do the sequential pattern in which the outcomes are sampled (sampling strategy) and the way the sampled outcomes are integrated and processed to evaluate the prospects (decision strategy) act together to shape the choice? This paper presents two models that incorporate the potential interplay between both types of strategies into the sequential sampling framework. Each model assumes an evidence accumulation process representing a distinct decision strategy and is used to simulate choices between a safe and a risky prospect, while systematically varying sampling strategies. The simulations show that the potential interplay between both types of strategies can lead to systematically different choice patterns and can contribute to the as-if underweighting of rare outcomes pattern in decisions from experience, irrespective of the existence or absence of sampling error. The model-implied differences in the choice patterns translate to distinct signatures in the psychoeconomic functions of cumulative prospect theory.

@@

The model assumes that the sampling process starts at random with a stochastic process on one of the two prospects.
Specifically, it is assumed that an agent starts with equal probability to sample from one of the prospects, say $P^X$, and generates a sequence of sampled outcomes, denoted by the random variables $X_1, X_2,...$, where the mean over this sequence is computed or updated with each new sampled outcome.
The respective stochastic process $\{S^X_t\}_{t \in \mathbb{N}}$ is defined by
$$
S^X_t = \frac{1}{t} \sum_{k = 1}^t X_k, \quad k,t \in \mathbb{N}
\; ,
\tag{1}
$$
where $S^X_t$ is the mean across $t$ outcomes that were *successively* sampled from prospect $P^X$.
Each such stochastic process terminates as soon as an outcome is sampled from the other prospect, i.e., the stochastic process on prospect $P^X$ stops as soon as an outcome is sampled from prospect $P^Y$, and vice versa.
The switching probability $\psi \in (0,1]$, which is assumed to be adopted prior the start of the sampling and accumulation process, is fixed throughout and controls the probability with which outcomes from different prospects are sampled in direct succession.
In other words, the switching probability $\psi$ governs the probability with which the stochastic process on a given prospect, defined as in (1), stops and a new stochastic process on the other prospect starts with the subsequently sampled outcome.

@@

The model assumes that such comparison rounds are repeatedly carried out until the cumulative sum of round wins for one of the two prospects reaches a threshold $\theta \in \mathbb{N}$.
Similar to the switching probability, it is assumed that the threshold $\theta$ is adopted prior the start of the sampling and accumulation process and is fixed throughout.
For $P^X$, then, one obtains a sequence of mean comparisons, denoted as $Z^X_1, Z^X_2, ...$, and an evidence accumulation process taking the form of a random walk $\{D^X_v\}_{v \in \mathbb{N}}$ defined by
$$
D^X_v = \sum_{c = 1}^v Z^X_c, \quad c,v \in \mathbb{N}
\;.
\tag{3}
$$
$D^X_v$ is the cumulative sum of round wins of prospect $P^X$ after $v$ comparison rounds, and resembles a dynamic decision variable.
Note that since (2) describes a Bernoulli trial, the random walk in (3) is based on a Bernoulli process.
Prospect $P^X$ would be chosen once $D^Y_v < \theta \leq D^X_v$ is satisfied, and vice versa.

@@

To implement the evidence accumulation process, the summary integration model assumes that an agent continues sampling from both prospects until the cumulative sum across all outcomes sampled from one of the prospects reaches a threshold $\theta$. 
As in the round-wise integration model, the threshold $\theta$ is assumed to be adopted prior the start of the sampling and accumulation process and is fixed throughout.
For $P^X$, then, one obtains a sequence of sampled outcomes, $X_1, X_2, ...$, and an evidence accumulation process taking the form of a random walk $\{D^X_t\}_{t \in \mathbb{N}}$ defined by 
$$
D^X_t = \sum_{k = 1}^t X_k, \quad k,t \in \mathbb{N}
\; ,
\tag{4}
$$
where $D^X_t$ is the cumulative sum across $t$ outcomes sampled from prospect $P^X$ and resembles a dynamic decision variable.
Assuming that $D^Y_m$ is the cumulative sum across $m$ sampled outcomes from prospect $P^Y$, prospect $P^X$ would be chosen once $D^Y_m < \theta \leq D^X_t$ is satisfied, and vice versa.

Note again that the described model assumes two random walks---i.e., one for each prospect---approaching the same threshold, which now is the cumulative sum across all sampled outcomes a prospect must produce in order to be chosen.
However, one may also define a single random walk, resembling a common dynamic decision variable, which approaches a positive threshold $\theta^+$ and a negative threshold $\theta^-$.
That is, (4) may be substituted by
$$
D^X_{t,m} = \sum_{k = 1}^t X_k - \sum_{j = 1}^m Y_j \quad k,t,j,m \in \mathbb{N}
\; ,
\tag{5}
$$
where $D^X_{t,m}$ is the difference between the cumulative sums across $t$ and $m$ outcomes sampled from prospects $P^X$ and $P^Y$, respectively.
The thresholds thus determine how much larger the cumulative sum of a prospect must be in order for the prospect to be chosen.

>>>>

authornote: |
 This manuscript is a dynamic document created from the materials provided on the `GitHub` repository: https://github.com/linushof/sampling-in-dfe. The current version of the manuscript, created from the commit with the hash ``r repro::current_hash()``, is submitted as a master's thesis in fulfillment of the requirements for the degree Master of Science (M.Sc.) at Heidelberg University, Faculty of Behavioural and Cultural Studies, Institute of Psychology. Matr.-Nr.: Hidden in online version. Supervisors:

@@

abstract: |
  In the sampling paradigm, people can---as much as they want, in the way they want, and without costs---sample the possible outcomes of prospects before making a final choice. Do the sequential pattern in which the outcomes are sampled (sampling strategy) and the way the sampled outcomes are integrated and processed to evaluate the prospects (decision strategy) act together to shape the choice? This paper presents two models that incorporate the potential interplay between both types of strategies into the computational framework of sequential sampling models. Each model assumes an evidence accumulation process representing a distinct decision strategy and is used to simulate choices between a safe and a risky prospect, while systematically varying sampling strategies. The simulations show that the potential interplay between both types of strategies can lead to systematically different choice patterns and can contribute to the as-if underweighting of rare outcomes pattern in decisions from experience, irrespective of the existence or absence of sampling error. The model-implied differences in the choice patterns translate to distinct signatures in the psychoeconomic functions of cumulative prospect theory.

@@

The model assumes that the sampling process starts at random with a stochastic process on one of the two prospects.
Specifically, it is assumed that an agent starts with equal probability to sample from one of the prospects, say $P^X$, and generates a sequence of sampled outcomes, denoted by the random variables $X_1, X_2,...$, where the mean over this sequence is computed or updated with each new sampled outcome.
The respective stochastic process $\{S^X_t\}_{t \in \mathbb{N}}$ is defined by
$$
S^X_t = \frac{1}{t} \sum_{k = 1}^t X_k, \quad k \in \mathbb{N}
\; ,
\tag{1}
$$
where $S^X_t$ is the mean across $t$ outcomes that were *successively* sampled from prospect $P^X$.
Each such stochastic process terminates as soon as an outcome is sampled from the other prospect, i.e., the stochastic process on prospect $P^X$ stops as soon as an outcome is sampled from prospect $P^Y$, and vice versa.
The switching probability $\psi \in (0,1]$, which is assumed to be adopted prior the start of the sampling and accumulation process, is fixed throughout and controls the probability with which outcomes from different prospects are sampled in direct succession.
In other words, the switching probability $\psi$ governs the probability with which the stochastic process on a given prospect, defined as in (1), stops and a new stochastic process on the other prospect starts with the subsequently sampled outcome.

@@

The model assumes that such comparison rounds are repeatedly carried out until the cumulative sum of round wins for one of the two prospects reaches a threshold $\theta \in \mathbb{N}$.
Similar to the switching probability, it is assumed that the threshold $\theta$ is adopted prior the start of the sampling and accumulation process and is fixed throughout.
For $P^X$, then, one obtains a sequence of mean comparisons, denoted as $Z^X_1, Z^X_2, ...$, and an evidence accumulation process taking the form of a random walk $\{D^X_v\}_{v \in \mathbb{N}}$ defined by
$$
D^X_v = \sum_{c = 1}^v Z^X_c, \quad c \in \mathbb{N}
\;.
\tag{3}
$$
$D^X_v$ is the cumulative sum of round wins of prospect $P^X$ after $v$ comparison rounds, and resembles a dynamic decision variable.
Note that since (2) describes a Bernoulli trial, the random walk in (3) is based on a Bernoulli process.
Prospect $P^X$ would be chosen once $D^Y_v < \theta \leq D^X_v$ is satisfied, and vice versa.

@@

To implement the evidence accumulation process, the summary integration model assumes that an agent continues sampling from both prospects until the cumulative sum across all outcomes sampled from one of the prospects reaches a threshold $\theta$. 
As in the round-wise integration model, the threshold $\theta$ is assumed to be adopted prior the start of the sampling and accumulation process and is fixed throughout.
For $P^X$, then, one obtains a sequence of sampled outcomes, $X_1, X_2, ...$, and an evidence accumulation process taking the form of a random walk $\{D^X_t\}_{t \in \mathbb{N}}$ defined by 
$$
D^X_t = \sum_{k = 1}^t X_k, \quad k \in \mathbb{N}
\; ,
\tag{4}
$$
where $D^X_t$ is the cumulative sum across $t$ outcomes sampled from prospect $P^X$ and resembles a dynamic decision variable.
Assuming that $D^Y_m$ is the cumulative sum across $m$ sampled outcomes from prospect $P^Y$, prospect $P^X$ would be chosen once $D^Y_m < \theta \leq D^X_t$ is satisfied, and vice versa.

Note again that the described model assumes two random walks---i.e., one for each prospect---approaching the same threshold, which now is the cumulative sum across all sampled outcomes a prospect must produce in order to be chosen.
However, one may also define a single random walk, resembling a common dynamic decision variable, which approaches a positive threshold $\theta^+$ and a negative threshold $\theta^-$.
That is, (4) may be substituted by
$$
D^X_{t,m} = \sum_{k = 1}^t X_k - \sum_{j = 1}^m Y_j \quad k,j \in \mathbb{N}
\; ,
\tag{5}
$$
where $D^X_{t,m}$ is the difference between the cumulative sums across $t$ and $m$ outcomes sampled from prospects $P^X$ and $P^Y$, respectively.
The thresholds thus determine how much larger the cumulative sum of a prospect must be in order for the prospect to be chosen.
